\documentclass[sigconf, nonacm]{acmart}
\IfFileExists{glyphtounicode.tex}{\input{glyphtounicode}\pdfgentounicode=1}{}
\usepackage[T1]{fontenc}
\usepackage{booktabs}
\usepackage{array}
\usepackage{float}
\usepackage{graphicx}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,fit,shapes.geometric,calc,backgrounds}
\usepackage{placeins}
\usepackage{amsmath,amsthm}
\usepackage{stmaryrd}
\usepackage{enumitem}
\Urlmuskip=0mu plus 2mu

\definecolor{vldbInk}{RGB}{28,31,35}
\definecolor{vldbBlue}{RGB}{38,86,132}
\definecolor{vldbTeal}{RGB}{37,124,116}
\definecolor{vldbGold}{RGB}{154,111,43}
\definecolor{vldbRose}{RGB}{151,73,82}
\definecolor{vldbViolet}{RGB}{87,82,141}
\definecolor{vldbMist}{RGB}{239,245,248}
\definecolor{vldbMint}{RGB}{237,247,244}
\definecolor{vldbWarm}{RGB}{248,244,235}
\definecolor{vldbRule}{RGB}{190,197,201}
\definecolor{vldbGray}{RGB}{96,105,113}
\definecolor{vldbSoft}{RGB}{246,248,250}

\tikzset{
  vtitle/.style={font=\rmfamily\bfseries\fontsize{8.8}{9.6}\selectfont,
    text=vldbInk, align=center, inner sep=.4pt, text height=1.65ex,
    text depth=.30ex},
  vhead/.style={font=\rmfamily\bfseries\fontsize{7.4}{8.2}\selectfont,
    text=vldbInk, align=center, inner sep=.5pt, text height=1.55ex,
    text depth=.30ex},
  vbody/.style={font=\rmfamily\fontsize{6.9}{7.6}\selectfont,
    text=black!68, align=center, inner xsep=2.0pt, inner ysep=1.0pt,
    text height=1.50ex, text depth=.28ex},
  vsmall/.style={font=\rmfamily\fontsize{6.7}{7.35}\selectfont,
    text=black!60, align=center, inner sep=.4pt, text height=1.45ex,
    text depth=.25ex},
  vpanel/.style={rounded corners=2.5pt, line width=.52pt},
  vflow/.style={-{Latex[length=1.8mm]}, line width=.74pt,
    draw=black!52, line cap=round, line join=round}
}

\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\newtheorem*{observation}{Observation}


\newcommand\vldbdoi{XX.XX/XXX.XX}
\newcommand\vldbpages{XXX-XXX}
\newcommand\vldbvolume{20}
\newcommand\vldbissue{1}
\newcommand\vldbyear{2027}
\newcommand\vldbauthors{\authors}
\newcommand\vldbtitle{\shorttitle}
\newcommand\vldbavailabilityurl{https://doi.org/10.5281/zenodo.21251990}
\newcommand\vldbarchivesha{\texttt{ac21995f\allowbreak{}c19c6c41\allowbreak{}086b5286\allowbreak{}fc4cdb70\allowbreak{}9408aec4\allowbreak{}fb424304\allowbreak{}519b3001\allowbreak{}eb5a4531}}
\newcommand\vldbpagestyle{plain}

\title{OptSem-C: Auditing Federated SQL Optimizer Contracts}

\author{Haoyi Zhang}
\orcid{0009-0009-3693-786X}
\affiliation{%
  \institution{Xi'an Jiaotong-Liverpool University}
  \city{Suzhou}
  \country{China}
}
\email{hyeliozhang@gmail.com}

\author{Huaijin Ran}
\orcid{0009-0009-2482-2344}
\affiliation{%
  \institution{Nanyang Technological University}
  \city{Singapore}
  \country{Singapore}
}
\email{huaijin003@e.ntu.edu.sg}

\author{Xunzhu Tang}
\affiliation{%
  \institution{University of Luxembourg}
  \city{Luxembourg}
  \country{Luxembourg}
}
\email{xunzhu.tang@uni.lu}

\author{Tegawend\'e F. Bissyand\'e}
\affiliation{%
  \institution{University of Luxembourg}
  \city{Luxembourg}
  \country{Luxembourg}
}
\email{tegawende.bissyande@uni.lu}

\AtBeginDocument{\hypersetup{%
  pdftitle={OptSem-C: Auditing Federated SQL Optimizer Contracts},
  pdfauthor={Haoyi Zhang, Huaijin Ran, Xunzhu Tang, and Tegawende F. Bissyande}}}

\begin{document}
\begin{abstract}
Database studies sometimes summarize optimizer features with compact labels such as pushdown, join, adaptive, or enabled.  In federated and cloud SQL systems, the same label can name different public contracts: a local rewrite, connector delegation, service-stage execution, runtime filtering, or only a plan-evidence surface.  A later latency, portability, or plan-quality claim can then compare the label rather than the optimizer behavior the public systems expose.  OptSem-C makes this denominator check executable.  It encodes each source-supported optimizer statement as a guarded public contract over action, guard, layer, placement, decision time, observability, and evidence support, then tests whether a declared comparison vocabulary collapses contracts that differ under the full public evidence payload.  In a seven-engine case study with 287 source-linked rules, 4,216 executable SQL probes, and 26 admitted public surfaces, keyword and operator-only vocabularies collapse 254 and 238 unequal public-contract signatures.  The failures are concentrated, not pervasive: they occur in four architectural-boundary pairs involving cloud-service, connector, and distributed-runtime behavior, while the other 17 engine pairs have no headline collision in this corpus.  The audit also gives a positive result: yes/no enablement controls add only six boundary witnesses.  Within the declared corpus, placement separates keyword witnesses, optimizer layer separates operator-only witnesses, and layer plus placement separates all 498 counted records.  OptSem-C is therefore a replayable finite audit, not a universal field ontology, for deciding when a comparison vocabulary is specific enough before downstream performance, portability, or plan-evidence claims are interpreted.
\end{abstract}

\maketitle

%%% do not modify the following VLDB block %%
%%% VLDB block start %%%
\pagestyle{\vldbpagestyle}
\begingroup\small\noindent\raggedright\textbf{PVLDB Reference Format:}\\
\vldbauthors. \vldbtitle. PVLDB, \vldbvolume(\vldbissue): \vldbpages, \vldbyear.\\
\href{https://doi.org/\vldbdoi}{doi:\vldbdoi}\par
\endgroup
\begingroup
\renewcommand\thefootnote{}\footnote{\noindent\raggedright\sloppy
This work is licensed under the Creative Commons BY-NC-ND 4.0 International License. Visit \url{https://creativecommons.org/licenses/by-nc-nd/4.0/} to view a copy of this license. For any use beyond those covered by this license, obtain permission by emailing \href{mailto:info@vldb.org}{info@vldb.org}. Copyright is held by the owner/author(s). Publication rights licensed to the VLDB Endowment. \\
\raggedright Proceedings of the VLDB Endowment, Vol. \vldbvolume, No. \vldbissue\ %
ISSN 2150-8097. \\
\href{https://doi.org/\vldbdoi}{doi:\vldbdoi} \\
}\addtocounter{footnote}{-1}\endgroup\par
%%% VLDB block end %%%
%%% do not modify the following VLDB block %%
%%% VLDB block start %%%
\ifdefempty{\vldbavailabilityurl}{}{%
\vspace{.3cm}
\noindent\begin{minipage}{\columnwidth}
\small\raggedright\textbf{PVLDB Artifact Availability:}\\
The source code, data, and/or other artifacts have been made available at \url{\vldbavailabilityurl}. Archive SHA-256: \vldbarchivesha.
\end{minipage}\par
}
%%% VLDB block end %%%

\input{sections/01_intro}
\input{sections/02_problem}
\input{sections/03_model}
\input{sections/04_benchmark_corpus}
\input{sections/05_evaluation}


\begingroup
\clubpenalty=10000
\widowpenalty=10000
\interlinepenalty=10000
\bibliographystyle{ACM-Reference-Format}
\bibliography{refs}
\endgroup
\end{document}
