\section{Introduction}
SQL deliberately hides physical execution choices.  That abstraction is right for applications, but a database study often needs to compare the hidden choice itself.  A benchmark author may ask whether two engines support pushdown, a system integrator may compare the evidence available in plan output, and a researcher may group systems by join, cost, or adaptivity labels before explaining a performance difference.  The same SQL query can then expose different public behavior: a local predicate rewrite, connector-side execution, a service-stage graph, a runtime filter, or only a logical explanation.  If a table states only ``pushdown=yes'', a later latency result may be comparing different public objects without saying so.

The problem is easiest to see in a concrete audited case.  Probe P0009 is a same-connector group-by aggregate over a remote customer table.  A keyword view equates BigQuery and Trino under the labels \texttt{pushdown} and \texttt{explain}.  The source records do not support the same public contract: BigQuery's federated record separates source delegation from Cloud SQL aggregate-pushdown blocking, Spanner partial aggregation, and stage-graph evidence, while Trino's pushdown record separates group-by aggregate pushdown from expression-aggregate blocking and estimated physical-plan evidence.  The practical rewrite is simple but consequential: a coarse statement such as ``both systems expose pushdown'' remains valid as an existence claim, while a portability, plan-evidence, or latency explanation must retain placement, layer, decision time, and observability.  Without those fields, the comparison has lost the denominator needed to decide whether two public claims were comparable in the first place.

This is not a claim that an existing paper necessarily reached a wrong conclusion.  It is a precondition check: before a paper attributes a difference to an optimizer feature, the compared feature has to be named at the level at which the public systems actually expose it.  OptSem-C therefore treats the vocabulary as a measured denominator, not as a survey of all optimizer papers.  We audit 26 public comparison surfaces drawn from engine documentation, plan interfaces, and optimizer-control descriptions admitted by the same source policy as the rules.  All expose keyword-style or operator-family labels, twelve expose yes/no-style controls, and none exposes the full reference payload used by OptSem-C.  The primary contribution is a replayable audit method plus an empirical boundary result: in this corpus, lossy labels fail at federated, cloud-service, and distributed-runtime boundaries, while most local-engine pairs and enablement-style controls do not create headline collisions.

Database research already has precise abstractions for the objects it chooses to study.  Relational semantics starts from Codd's model~\cite{codd1970relational}; Selinger-style optimizers separate access-path search from costing~\cite{selinger1979access}; Volcano exposes rule-driven and parallel query processing~\cite{graefe1990volcano}; the Volcano optimizer-generator account makes extensible search explicit~\cite{graefe1993volcano}; and Cascades turns that line into a task-driven framework~\cite{graefe1995cascades}.  Chaudhuri surveys optimizer architecture~\cite{chaudhuri1998overview}; Ioannidis surveys cost and cardinality assumptions~\cite{ioannidis1996query}; and Steinbrunn et al. study join-order search heuristics~\cite{steinbrunn1997heuristic}.  Public optimizer behavior is usually compared with much coarser vocabularies.  A cell labeled ``pushdown'' may mean predicate movement inside the engine, partition pruning, aggregation delegation, join delegation, limit delegation, or runtime filtering.  A cell labeled ``join'' may refer to a logical operator, an estimated physical operator, a distributed exchange, or a runtime profile entry.  A claim that two engines are ``cost based'' may hide whether statistics are local, connector-provided, unavailable, or revised during execution.  These shortcuts are not harmless summarization: they can create projection-induced contract collisions (a coarse vocabulary declares equality among public contracts whose reference signatures differ).

OptSem-C treats this precision problem as a query-processing systems problem, not as prose cleanup.  A \emph{public optimizer contract} is the optimizer behavior an engine exposes through plan interfaces, tuning surfaces, and system descriptions.  A source record can support a contract that is not visible in a particular textual plan, and a plan tag can expose behavior without proving the private cost model.  The object is therefore not a hidden implementation trace and not a latency measurement.  OptSem-C represents each public statement as a guarded contract rule over query features.  The rule records the action, the guard under which it applies, where the action is placed, which optimizer layer owns it, when the decision is made, how the behavior is observable, and whether the public contract requires, permits, forbids, or does not specify the action.

\textbf{Research questions.}  We ask four finite, auditable questions.  RQ1 asks whether public optimizer claims can be encoded as source-linked contracts and executable probes without becoming a latency benchmark.  RQ2 asks when the declared comparison vocabularies collapse distinct source-supported contracts inside the admitted surface catalog, and when coarse controls remain benign.  RQ3 asks which semantic fields separate those collisions, and which robustness checks delimit the corpus-specific certificate.  RQ4 asks whether the projection-audit inner loop and SQL-denominator checks are cheap enough to rerun when the public source denominator changes, rather than trusted as a one-off manual inspection.

The central result is a corpus-specific projection-kernel account of information loss in this seven-engine corpus.  Comparing optimizer contracts through a keyword or operator vocabulary applies a predeclared projection to reference signatures: the full public evidence payload inside the declared schema, not an oracle over private optimizer state.  If reference signatures disagree but fall into the same projected class, the projection has declared an equivalence that source-supported public signatures do not support.  The witness does not prove an optimizer bug or a false performance result.  It proves that an unqualified equality claim needs either a narrower scope statement or retained fields before downstream measurements can be interpreted.

OptSem-C turns that denominator question into a finite comparison relation with source-linked counter-witnesses, executable SQL representatives, exact separation certificates, and negative controls.  Figure~\ref{fig:contract-model} shows the end-to-end boundary.  The contract schema, source manifest, projection portfolio, feature-space crosswalk, and retained-field universe are frozen before witness counting; execution checks validate probes but never admit rules or choose fields.  Layer and placement are architecture hypotheses from optimizer and federation literature; the replay confirms that this predeclared field pair is the most interpretable basis among the finite safe certificates, not an empirical discovery procedure and not proof of a universal vocabulary.  The test is falsifiable: placement alone fails operator-only, layer alone leaves keyword witnesses, and point-learned engine-pair bases fail held-out pairs.  Keyword projection declares 2,284 engine--probe equivalences, including 254 collapsed unequal signatures; operator-only declares 2,268, including 238.  These 492 primary witnesses occupy four observed engine pairs and 486 distinct probes.  The binary yes/no projection contributes only six additional records, counter-evidence to a broad binary-control danger claim.  Retaining placement separates keyword witnesses; retaining optimizer layer separates operator-only witnesses; retaining layer and placement also covers the yes/no boundary records within the admitted public-contract corpus.

The robustness checks are designed to block three tempting overclaims.  First, the probe suite is a rule-aware coverage test, not an independent population sample, so we do not report statistical generalization beyond the frozen denominator.  Second, the four engine-pair hotspots are not hidden: they show that projection risk is concentrated around cloud-service, connector, and distributed-runtime boundaries rather than uniformly spread across all engine pairs.  Third, source-removal is an information-deletion stress rather than a new sample: keyword and operator-only remain nonzero after every single-source removal, and the audit separately records retained original witnesses and newly induced witnesses.  The sparse yes/no projection is source-sensitive and is reported as a narrow public-control result, not as evidence that every binary feature table is safe.  Last, field certificates are corpus certificates: the reported layer-and-placement basis resolves feature-family and engine-family stress splits, while point-learned engine-pair bases fail on held-out pairs and remain a boundary.  A field-only stress test over all 256 retained-field vocabularies leaves only 44 unsafe regions, all covered by concrete counter-witnesses; after excluding action-variant identifiers, no singleton semantic field is safe.

We also introduce OptSemBench-C, a contract-probe suite for optimizer-comparison audits.  Its denominator is not raw query count or workload runtime.  It is coverage of optimizer-relevant feature interactions: source boundary, connector capability, join shape, predicate class, aggregation, statistics need, reuse, distribution, adaptivity, and control surface.  The generated probes are executable SQL representatives.  The probe space is then cross-checked against DPccp-style join enumeration~\cite{moerkotte2006dpccp}, JOB optimizer evidence~\cite{leis2015good}, later JOB diagnostics~\cite{leis2018looking}, exact-cardinality testing~\cite{chaudhuri2009exact}, progressive optimization~\cite{markl2004progressive}, mid-query reoptimization~\cite{kabra1998reoptimization}, adaptive query processing~\cite{deshpande2007adaptive}, and eddies~\cite{avnur2000eddies}.  We use this audit as feature-space coverage of published optimizer requirements: 90 declared requirements are covered by 71 distinct generated SQL representatives.  The check is concrete about optimizer-feature span without pretending that generated representatives reproduce JOB, TPC-H, or TPC-DS workloads, data distributions, or selectivity effects.

\textbf{Contributions.} C1 defines a public optimizer contract as a query-processing object separate from SQL result equivalence, runtime performance, and private implementation truth; Section~\ref{sec:model} gives the guarded rule model and the state-join semantics.  C2 turns comparison validity into a finite projection audit: reference signatures define the public evidence relation, projected signatures expose counter-witnesses, and retained-field certificates identify fields that make the observed vocabulary specific enough.  C3 contributes OptSemBench-C as a contract-probe denominator for optimizer-comparison audits across seven analytical SQL engines; Section~\ref{sec:benchcorpus} reports source-linked admission, executable probes, feature-space coverage, and SQL-constructability checks.  C4 reports corpus-specific positive and negative projection results: 254 keyword witnesses, 238 operator-only witnesses, the six-record yes/no boundary, architecture-readable field bases for those finite records, and cloud replay evidence for the comparison inner loop.

\input{figures/fig_contract_model}

Each central claim is tied to a reproducible record rather than to prose alone.  The model section defines the state table and projection obligations; the corpus section exposes source identifiers, line spans, and digest prefixes; the evaluation tables connect SQL execution, feature-space coverage, collision counts, and retained-field certificates to regenerated outputs.  The case studies then show how those aggregate certificates change comparison judgments.
