\section{Precision Loss in Optimizer Comparisons}
\label{sec:problem}
Figure~\ref{fig:projection-trap} gives the smallest example.  A feature matrix maps several public optimizer contracts into the same label.  After the mapping, behaviors with different placement, action kind, and evidence layer become indistinguishable, so a study that treats the label as equality can compare non-equivalent public optimizer behavior.  Existing benchmarks and optimizer-comparison methods standardize workloads, cardinalities, or plan-quality measurements; they do not define the equality relation over public pushdown, adaptation, or plan-evidence contracts.  OptSem-C fills that gap before a JOB-, TPC-, or engine-specific measurement is interpreted.

\input{figures/fig_projection_trap}

\textbf{Representative witness audits.}  Table~\ref{tab:cases} gives six concrete witness rows before the formal model.  Each starts from a familiar public label, names the engine pair and probe, shows the contract distinction that the label hides, and gives the comparison rewrite.  C1 is the running P0009 example; C2--C3 cover keyword join/plan evidence, C4--C5 cover operator-only plan and join rows, and C6 is the narrow yes/no boundary.  The same finite checker covers the 492 primary keyword/operator records and the six yes/no boundary records.

\input{tables/tab_cases}

C1 is mechanical.  Source spans admit rules; P0009 triggers them with a remote aggregate; reference signatures differ on placement, layer, decision time, and observability; and the keyword projection maps both to \texttt{pushdown}+\texttt{explain}.  A coarse existence claim may remain, but portability or latency explanations must keep those fields.

\textbf{Feature-name collision.} A cell such as \texttt{pushdown=yes} may denote predicate movement inside an engine, partition pruning, connector-side predicate delegation, aggregation delegation, join delegation, limit pushdown, Top-N delegation, or runtime filtering.  These actions affect different operators and have different guards, placements, decision times, and observable plan effects.  Runtime and adaptivity distinctions are visible in Tukwila's adaptive integration~\cite{ives1999tukwila}, dynamic query plans~\cite{cole1994optimization}, and Spark SQL adaptive execution~\cite{armbrust2015spark}.  Rewrite and execution-layer distinctions appear in outerjoin rewrites~\cite{galindo1997outerjoin}, compiled execution~\cite{neumann2011efficient}, Calcite's adapter-centered optimizer framework~\cite{begoli2018calcite}, Orca's modular optimizer architecture~\cite{soliman2014orca}, and Dremel-style serving trees~\cite{melnik2010dremel}.  A single feature name can therefore erase exactly the fields a comparison needs.

\textbf{Operator disappearance.} A missing native plan node is not a reliable semantics.  It may indicate source delegation, logical-plan observability, rewrite, stage-level aggregation, or runtime-only visibility.  OptSem-C represents source-delegated execution as a positive contract action rather than as absence.  This matters for modern analytical systems whose public surfaces combine Snowflake-style cloud-service stages~\cite{dageville2016snowflake}, Redshift's managed warehouse architecture~\cite{gupta2015redshift}, C-Store column layout~\cite{stonebraker2005cstore}, row/column execution differences~\cite{abadi2008columnstores}, column-oriented design guidance~\cite{abadi2013columnoriented}, materialization policy~\cite{abadi2007materialization}, compression-coupled execution~\cite{abadi2006compression}, embedded DuckDB execution~\cite{raasveldt2019duckdb}, vectorized MonetDB/X100 execution~\cite{boncz2005monetdb}, HyPer's compiled in-memory engine~\cite{kemper2011hyper}, Vectorwise vectorized execution~\cite{zukowski2012vectorwise}, Hive's MapReduce warehouse layer~\cite{thusoo2009hive}, Impala's Hadoop SQL engine~\cite{kornacker2015impala}, distributed SQL services such as F1~\cite{shute2013f1}, and dataflow-style execution such as Spark RDDs~\cite{zaharia2012resilient}.

\textbf{Architecture ranking.} A local engine should not be ranked below a distributed engine merely because broadcast, shuffle, or remote-source delegation is not applicable.  Conversely, a federated or cloud engine should not be credited with local optimizer behavior when the public contract exposes delegation or service-side stages.  OptSem-C separates non-applicability, documented absence, documented optional behavior, and unspecified evidence.

\textbf{Observable contract versus hidden implementation.} The object of study is not what an engine could do internally.  It is what the public contract supports as a comparison claim.  This distinction is analogous to the separation between SQL equivalence and optimizer-contract equivalence.  Cosette proves SQL equivalences by reducing query semantics to logic~\cite{chu2017cosette}; semiring-based decision procedures give another equivalence foundation~\cite{chu2018usemiring}; and QED expands practical SQL equivalence checking~\cite{wang2024qed}.  Chandra and Merlin characterize conjunctive-query containment~\cite{chandra1977optimal}; Aho et al. characterize equivalence among relational expressions~\cite{aho1979equivalences}.  OptSem-C asks a different question: whether the public optimizer behavior being compared exposes the same evidence layers and comparison obligations.

These precision losses imply a minimum payload for comparing public optimizer behavior: guard, canonical action, modality, placement, decision time, observability, version, and provenance.  The rest of the paper formalizes this payload and measures what happens when a comparison vocabulary projects it away.
