\section{Contract Formalization and Notation}
\label{sec:model}
A query probe is a SQL skeleton paired with a finite feature vector and validity constraints.  Features describe optimizer-relevant structure: join shape and type, predicate class, aggregation, order/limit behavior, source boundary, connector capability, statistics need, reuse structure, distribution trigger, adaptivity trigger, and optimizer control surface.  They are not workload-frequency estimates; they are coordinates on the public optimizer decision surface.

The running P0009 witness is a concrete instance before the notation.  Under a coarse vocabulary, BigQuery and Trino can both appear as \texttt{pushdown}+\texttt{explain}.  The replay ledger binds the pair to BigQuery federated/stage sources and Trino pushdown sources, plus an executable remote \texttt{GROUP BY COUNT(*)} probe.  The contract payload keeps the distinctions that the label drops: BigQuery records source delegation, stage-graph observability, Cloud SQL aggregate-pushdown blocking, and Spanner partial aggregate pushdown, while Trino records connector aggregate pushdown, expression-aggregate blocking, and estimated physical-plan evidence.  The model below preserves those public distinctions without asserting a private optimizer trace.

The model is task relative.  A projection is safe only for the comparison question it declares.  If a study deliberately asks the narrow question ``does any public source expose a pushdown-like label?'', then a keyword projection may be adequate.  If the study compares plan evidence, connector delegation, or runtime behavior, then the same projection can be under-specified.  OptSem-C reports this under-specification as a collapsed public-contract comparison, not as an optimizer correctness error.

The retained fields are also task relative.  OptSem-C does not learn a field ontology from collisions.  It starts from a predeclared architecture vocabulary drawn from optimizer and federation literature, then asks which retained-field subsets separate the finite witness set.  The result is therefore confirmation of a stated ontology on this corpus, not discovery of the only correct vocabulary.

\begin{definition}[Public contract rule]
Let $\mathcal{A}$ be the finite domain of canonical optimizer actions.  An action records operator class, action kind, optimizer layer, execution placement, decision time, observability, and a variant name for source-specific subcases such as aggregate-expression blocking or partial aggregation.  The state set $\mathbb{S}$ contains $\mathsf{MUST}$, $\mathsf{MAY}$, $\mathsf{MUST\_NOT}$, and $\mathsf{UNSPEC}$.  A public contract rule is $r=(\gamma,a,s,\nu,\epsilon)$, where $\gamma$ is a finite disjunction of feature literals over $\phi(q)$, $a\in\mathcal{A}$, $s\in\mathbb{S}$, $\nu$ is the version or time window, and $\epsilon$ is evidence provenance.  For engine $e$, $C_e$ is the finite rule set, and $C_e(q)=\{r\in C_e\mid \gamma_r(\phi(q))=\mathsf{true}\}$ is the rule set applicable to probe $q$.
\end{definition}

The state domain has two components that feature matrices usually conflate.  Behaviorally, $\mathsf{MUST}$ requires an action, $\mathsf{MUST\_NOT}$ forbids it, and both $\mathsf{MAY}$ and $\mathsf{UNSPEC}$ leave it possible.  Evidentially, $\mathsf{MAY}$ is public support for optional behavior, while $\mathsf{UNSPEC}$ is lack of public support.  Equivalently, a state is interpreted as $(m,u,e)$: required behavior, allowed behavior, and evidence support.  This is why documented optional pushdown and undocumented implementation possibility are different public contracts.

For each engine-probe pair, OptSem-C constructs a contract map $M_{e,q}:\mathcal{A}\rightarrow\mathbb{S}$, defaulting every action to $\mathsf{UNSPEC}$.  Applicable rules update the map with a conflict-aware state join $\sqcup$.  The join strengthens unspecified actions when public evidence is found, preserves compatible stronger evidence, and reports contradictions instead of averaging them.  The evidence component follows database-provenance practice: semiring provenance attaches support to derived facts~\cite{green2007provenance}, why/where provenance distinguishes explanation locations~\cite{buneman2001why}, provenance surveys separate data from explanation~\cite{cheney2009provenance}, and annotated containment keeps evidence annotations visible in reasoning~\cite{green2011containment}.  The workflow in Figure~\ref{fig:contract-model} materializes these maps, tests projected equality against reference inequality, and records retained-field certificates.  Table~\ref{tab:state-join} gives the local algebra used inside each map.

\input{tables/tab_state_join}

\begin{definition}[State-join algebra]
The binary operator $\sqcup$ is the complete $4\times4$ partial join shown in Table~\ref{tab:state-join}.  It is total over non-conflicting pairs and returns $\mathsf{CONFLICT}$ for pairs that simultaneously require and forbid, or support and forbid, the same canonical action.  Let $x\preceq y$ denote that $x\sqcup y=y$ for non-conflicting states.  Then $\mathsf{UNSPEC}$ is the least informative state, while $\mathsf{MUST}$ and $\mathsf{MUST\_NOT}$ are incomparable incompatible commitments.
\end{definition}
Here $\mathsf{CONFLICT}$ is a merge-failure signal, not a fifth valid contract state.  A conflicting public record is split by guard/version when the sources name distinct conditions, or excluded when no public condition separates the statements.

\begin{lemma}[Merge invariants]
On non-conflicting states, $\sqcup$ is commutative, associative, idempotent, and monotone with respect to $\preceq$.  Therefore a contract map obtained by joining applicable rules is independent of rule order.
\end{lemma}
\begin{proof}
The properties follow by inspection of the finite table.  Idempotence holds because every diagonal entry is the input state.  Commutativity holds because the table is symmetric.  Associativity and monotonicity are checked over the finite non-conflicting triples; the only excluded triples are exactly those containing incompatible positive and negative commitments.  Lifting the pointwise operator to maps preserves the properties for every action independently.
\end{proof}

Let $W\subseteq\mathcal{A}$ be an abstract optimizer-behavior world.  A map $M$ denotes the world set
$\llbracket M\rrbracket=\{W\mid M(a)=\mathsf{MUST}\Rightarrow a\in W,\; M(a)=\mathsf{MUST\_NOT}\Rightarrow a\notin W\}$.
The evidential support set is $\mathsf{Supp}(M)=\{a\in\mathcal{A}\mid M(a)\in\{\mathsf{MUST},\mathsf{MAY},\mathsf{MUST\_NOT}\}\}$, equivalently the actions where $M(a)\ne\mathsf{UNSPEC}$.  Since maps default to $\mathsf{UNSPEC}$, every supported non-default entry arises from an admitted public rule.  Behavioral containment and evidential support are intentionally separate: two maps can allow the same worlds while differing on whether an action is publicly documented.

\begin{definition}[Reference signatures, kernels, and projections]
The reference evidence-supported signature of $M$ is $S(M)=\{(a,M(a))\mid a\in\mathsf{Supp}(M)\}$.  Reference equivalence is equality of signatures over the full public evidence atom schema.  This reference is exact only relative to the admitted public evidence and declared atom fields; it is not a complete private optimizer trace.  A projection is a declared map $\pi:\mathcal{A}\times\mathbb{S}\rightarrow\mathcal{B}$ from full evidence atoms to an arbitrary comparison vocabulary $\mathcal{B}$, such as keyword, yes/no, or operator-only labels.  The projected signature is the finite image $S_\pi(M)=\{\pi(a,M(a))\mid a\in\mathsf{Supp}(M)\}$.  The projection kernel is the equivalence relation $\mathcal{K}_\pi$ where $M_i\mathcal{K}_\pi M_j$ iff $S_\pi(M_i)=S_\pi(M_j)$.  A projection-induced contract collision is a pair in $\mathcal{K}_\pi$ whose reference signatures differ.
\end{definition}

The reference signature is deliberately stronger than many comparison tasks.  Thus a collision is a warning about an unqualified equality at projection $\pi$: the paper or tool using that vocabulary should either accept the coarser task, state the lost distinctions as out of scope, or retain fields that separate the witness.

For quantitative comparison, exact compatibility is the Jaccard overlap $|S(M_i)\cap S(M_j)|/|S(M_i)\cup S(M_j)|$, and projected compatibility applies the same formula to $S_\pi$.  The denominator is evidence-supported: undocumented actions do not count as evidence of either equality or inequality.

\begin{lemma}[Public-contract separation]
\label{thm:public-contract-separation}
If $S(M_i)\ne S(M_j)$, then the maps differ in public-contract denotation: either $\llbracket M_i\rrbracket\ne\llbracket M_j\rrbracket$, or their evidence-supported action states differ for at least one action in $\mathsf{Supp}(M_i)\cup\mathsf{Supp}(M_j)$.
\end{lemma}
\begin{proof}
If reference signatures differ, some supported action has a state in one map that is absent or different in the other.  If the difference changes a $\mathsf{MUST}$ or $\mathsf{MUST\_NOT}$ requirement, the world-set constraints differ.  Otherwise the behavioral world set may remain equal, but the public evidence state differs, for example $\mathsf{MAY}$ versus $\mathsf{UNSPEC}$.  In both cases the denotation differs because an OptSem-C contract is the pair of behavioral worlds and evidence-supported states.
\end{proof}

For finite feature and action domains, constructing $M_{e,q}$ from finite $C_e$ and $q$ is a direct finite evaluation: each guard is a Boolean formula over the probe feature vector, and each applicable rule performs one lookup in the state-join table.

\begin{observation}[Projection-kernel collision]
\label{thm:projection-kernel-loss}
For any projection $\pi$, a projection-induced contract collision is exactly the nontrivial part of the kernel $\mathcal{K}_\pi$: $M_i$ and $M_j$ collide iff $S(M_i)\ne S(M_j)$ and $M_i\mathcal{K}_\pi M_j$.  If $\pi$ is non-injective on evidence atoms, there exist contracts with such a collision witness.
\end{observation}
\begin{proof}
The first statement unfolds the definitions: $\mathcal{K}_\pi$ is equality after projection, while exact non-equivalence is $S(M_i)\ne S(M_j)$.  For the non-injective case, take distinct evidence atoms $(a_1,s_1)\ne(a_2,s_2)$ with $\pi(a_1,s_1)=\pi(a_2,s_2)$.  Construct $M_1$ so that $S(M_1)=\{(a_1,s_1)\}$ and $M_2$ so that $S(M_2)=\{(a_2,s_2)\}$, with all other actions $\mathsf{UNSPEC}$.  These minimal contracts are well formed because action states are assigned independently.  Then $S(M_1)\ne S(M_2)$, but $S_\pi(M_1)=S_\pi(M_2)$ is the same singleton, so $(M_1,M_2)$ is a collision witness.
\end{proof}

\begin{definition}[Projection information profile]
For a finite set of observed maps $Y$, let $H(S)$ be the empirical Shannon entropy of reference signatures and $H(S_\pi)$ the entropy of projected signatures.  The entropy-retention ratio $\rho_\pi=H(S_\pi)/H(S)$ measures distinguishability retained by the comparison vocabulary.  The projected collision count is the number of projected-signature classes containing more than one reference signature class.
\end{definition}

\begin{lemma}[Kernel inflation]
If a projection declares $E_\pi$ equivalent pairs and reference comparison declares $E$ over the same pair set, then $E_\pi/E$ is the pairwise inflation factor induced by the kernel.  Every collision witness contributes to this inflation, and a zero-collision strengthened projection must have $E_\pi=E$.
\end{lemma}
\begin{proof}
Reference equivalence implies projected equivalence only when the projection is a function of the full atom payload, so $E_\pi\ge E$.  The additional pairs counted by $E_\pi$ are exactly the unequal reference signatures that collapse under $\pi$, which are collision witnesses.  If there are none, no additional pairs are counted and the two denominators agree.
\end{proof}

\begin{observation}[Evidence refinement]
Adding a non-conflicting evidence-backed rule cannot silently weaken a public contract: it behaviorally refines possible worlds, evidentially refines support, or both.
\end{observation}

The remaining notation is the payoff of the model: it turns an informal vocabulary dispute into a finite certificate check.  Let $F$ be the semantic-field universe and let $\pi_R$ append field values in $R\subseteq F$ to the baseline projection.  For a finite comparison set $X$, define $X_{\pi_R}=\{(M_i,M_j)\in X\mid S(M_i)\ne S(M_j)\land M_i\mathcal{K}_{\pi_R}M_j\}$.  A retained-field set $R$ is safe for $\pi$ on $X$ when $X_{\pi_R}=\emptyset$; otherwise any pair in $X_{\pi_R}$ is a counter-witness.  Since adding retained fields can split projected classes but cannot merge already separated classes, safe sets are upward closed and unsafe sets are downward closed.  For each witness $w$, the separator family $\mathcal{D}_w=\{R\subseteq F\mid M_i\not\mathcal{K}_{\pi_R}M_j\}$ is upward closed, so a universal retained-field separator is the finite intersection problem $R\in\bigcap_{w\in X_\pi}\mathcal{D}_w$.  The reported certificates therefore consist of all separated witnesses plus counter-witnesses for smaller field sets; exhaustive enumeration over $2^{|F|}$ retained-field subsets is complete for the declared field universe.

\begin{observation}[Finite retained-field separator search]
\label{thm:min-separator-complexity}
For OptSem-C's fixed semantic-field universe, exhaustive retained-field enumeration is a complete certificate search for the declared corpus: every reported safe set has no remaining witness, and every smaller rejected set has a concrete counter-witness.
\end{observation}
\begin{proof}
Each retained-field set $R\subseteq F$ induces a finite projected signature for every observed map.  Direct evaluation therefore decides whether $X_{\pi_R}$ is empty.  Enumerating all $2^{|F|}$ subsets visits every possible retained-field vocabulary in the declared universe; the upward-closure of safe sets and the retained counter-witness rows explain why a smaller rejected set failed.  The certificate is complete for this finite universe, but it is not a claim that the same fields form a universal ontology outside the admitted corpus.
\end{proof}

The semantics makes the comparison payload explicit.  A user comparing only operator names chooses a projection that drops placement, decision time, variant, and observability.  A feature checklist chooses an even coarser projection that can drop operator class itself.  OptSem-C does not forbid those projections; it computes exactly which kernel collisions they induce, which grounded counter-witnesses remain, and which query-processing fields separate the comparison within the declared corpus.
