\section{Contract Probe Suite and Grounded Corpus}
\label{sec:benchcorpus}
OptSemBench-C is a contract-probe suite: it generates executable SQL representatives that cover optimizer-relevant feature interactions and measures which public contracts those probes activate.  It deliberately does not rank engines by latency.  This choice follows the measurement principle that the metric and denominator must match the object of study: large-scale system comparisons measure analytical designs~\cite{pavlo2009comparison}, while OptSemBench-C measures whether the comparison vocabulary preserves public optimizer distinctions.  Cardinality benchmarks ask whether estimates improve optimizer decisions, as in CardEst~\cite{han2021cardest} and CardBench~\cite{chronis2024cardbench}.  Learned estimators such as Learned Cardinalities~\cite{kipf2019learned}, NeuroCard~\cite{yang2019neurocard}, DeepDB~\cite{hilprecht2020deepdb}, and FactorJoin~\cite{wu2023factorjoin} model distributions or joins; learned optimizers such as Neo~\cite{marcus2019neo} and Bao~\cite{marcus2021bao} test plan selection or hinting.  OptSemBench-C asks a prior question: before measuring plan quality, did the comparison vocabulary preserve the public optimizer distinctions being compared?

The contract schema is fixed before source extraction, witness counting, and retained-field search.  Its fields come from prior optimizer architecture rather than from the measured collisions.  Operator/action and layer follow System R~\cite{selinger1979access}, Volcano~\cite{graefe1990volcano}, Cascades~\cite{graefe1995cascades}, Calcite~\cite{begoli2018calcite}, and Orca~\cite{soliman2014orca}.  Placement follows Garlic federation~\cite{carey1995garlic}, BigDAWG polystores~\cite{stonebraker2015bigdawg}, Dremel serving trees~\cite{melnik2010dremel}, and Snowflake cloud execution~\cite{dageville2016snowflake}.  Decision time follows progressive optimization~\cite{markl2004progressive}, mid-query reoptimization~\cite{kabra1998reoptimization}, adaptive query processing~\cite{deshpande2007adaptive}, eddies~\cite{avnur2000eddies}, and Spark adaptive execution~\cite{armbrust2015spark}.  Observability is admitted from plan, profile, and stage evidence in the public source ledger, with provenance-style evidence separation~\cite{green2007provenance}.  Layer and placement enter the field universe as architecture-derived dimensions; exhaustive retained-field testing later reports them as an interpretable sufficient basis among finite certificates, not as a universal separator learned from private behavior.

Let $F=F_1\times\cdots\times F_n$ be the finite feature domain and $\Psi$ the validity constraints.  A target interaction is a valid partial assignment over a feature subset.  The schema and feature domain are fixed before witness counting and derive from standard query-processing constructs and cited public optimizer surfaces, not from witnesses or motifs.  The probe suite targets global pairwise coverage plus selected three-way interactions for source boundary, connector capability, operator class, statistics need, join shape, adaptivity trigger, and distribution trigger.  The generator builds one deterministic candidate pool from renderable feature interactions and admitted public-rule guards.  It first includes 99 guard-forced completions so that every admitted rule has a reachable query shape, then greedily covers the remaining feature-interaction universe.  These two roles are reported separately: guard-forced probes prevent public contracts from becoming unreachable paper claims, while the remaining probes complete the declared feature-interaction denominator.  Neither role is treated as an independent held-out population sample.  A probe is useful when it activates a contract guard or separates two contract signatures, not when it estimates workload frequency.

\begin{observation}[Coverage]
If every valid target interaction has at least one renderable full assignment, the generator terminates with a probe set covering every renderable target interaction.
\end{observation}
\begin{proof}
The target universe is finite.  Each selected assignment covers at least one uncovered interaction.  Thus the number of uncovered interactions decreases monotonically until it reaches zero.
\end{proof}

The main empirical corpus contains only 287 source-linked grounded rules.  Each rule records a public evidence span, source URL, line range, digest, and source class.  Claims without grounded evidence, expressible guards, or public observability are outside the finite denominator rather than negative evidence.  Thus the paper claims coverage of the declared public-contract denominator, not a census of private optimizer rules.  Figure~\ref{fig:benchmark-snapshot} gives the denominator view before any collision measurement: sources, rules, probes, and feature-space requirements are fixed, and the zero-failure checks verify support, consistency, SQL execution, and crosswalk coverage.  The corpus covers plan observability, pushdown, join search, distribution, materialization, runtime adaptivity, statistics, and optimizer control surfaces.  Every grounded rule is triggered by at least one generated probe.  Zero admitted merge conflicts means unresolved MUST/MUST\_NOT evidence has been split by guard/version or excluded before admission, not that public documentation never conflicts.

\input{figures/fig_benchmark_snapshot}

\input{tables/tab_evidence_examples}

The source set is fixed before witness counting.  We admit an engine only when public sources provide: a comparable plan or contract surface, enough detail to extract an action and applicability guard, and stable analytical-SQL behavior rather than a private vendor note.  PostgreSQL, Trino, BigQuery, Spark SQL, Snowflake, DuckDB, and ClickHouse enter through versioned source records in the archived evidence ledger named by the page-1 availability DOI and SHA; Calcite and DataFusion source records provide schema context rather than denominator engines.  These records are not scholarly support and not performance measurements.  They are the measured public contracts that the framework formalizes.  Proprietary workload logs, undocumented cost-model traces, future engine versions, and operator families outside the declared feature domain are excluded rather than silently counted.  This admission rule makes documentation density a visible confound: source-rich engines can create more measured collision opportunities, so the paper reports finite source sensitivity rather than representativeness.

The source-linked denominator spans seven engines: PostgreSQL contributes 80 admitted rules, Trino 61, BigQuery 37, Spark SQL 36, ClickHouse and Snowflake 26 each, and DuckDB 21; the largest single source supports 46 rules.  This design separates admission, probing, and execution validation.  Public sources admit rules; generated probes test guard reachability and SQL executability; DuckDB and PostgreSQL runs validate the probe denominator on two open engines, not every vendor contract.  All-engine execution would conflate SQL validity with public-contract admission.  A DuckDB source span can admit a DuckDB public contract, but a later DuckDB execution is only SQL execution and visible-plan sanity evidence; it never admits rules or chooses a separating field.  Conversely, engines without local execution still participate in collision measurements only through source-linked public contracts, source leave-one-out checks, and side-balanced witness support.

The grounded corpus is intentionally conservative.  A rule is admitted only when the evidence span supports an optimizer action, a guard or applicability condition, and an observability or placement interpretation.  Vague performance advice is excluded from the main corpus.  This policy reduces corpus size but strengthens the central claim: the evaluation is about contracts that are publicly supported, not about inferred implementation behavior.

\textbf{Contract extraction and validation checks.}  Each admitted rule records the engine, version scope, source identifier, digest, action fields, guard, and observability surface.  The admission codebook is operational rather than statistical: a span must name an optimizer action, state an applicability condition or guard, expose observability or execution placement, and map to a versioned public locator; otherwise it is excluded from the denominator rather than converted into negative evidence.  Ambiguous, advisory-only, or performance-only prose is excluded by the same rule.  Public contracts are therefore constructed interpretations of public evidence, not objective claims about hidden optimizer behavior.  Automated checks validate schema consistency, source linkage, guard reachability, trigger coverage, merge conflicts, SQL executability, and table-source consistency.  This is a single-extractor source-line protocol rather than formal verification or recoding agreement, so the paper does not claim inter-rater reliability.  The reliability claim is instead auditability: every accepted row has a locator, line span, digest, guard, paraphrase, and regenerated downstream witness; every excluded row is outside the declared denominator rather than hidden as a negative.  The public-provenance ledger contains 287 verified rules, 287 source segments, 26 source digests, retrieval dates, paraphrase policy, and zero public-provenance issues under replay checks; the table-source audit maps all paper tables to regenerated evidence files.  Table~\ref{tab:evidence-examples} gives representative spans; Table~\ref{tab:projection-surface-portfolio} exposes the source-surface catalog; the replay materials carry the complete rule-to-source manifest, source digests, claim-evidence graph, regenerated table manifests, per-row surface flags, and a recoding worksheet with blank independent-coding columns for independent follow-up.  This evidence chain lets an auditor recode or dispute any row without trusting hidden notes, but it remains a manual public-contract extraction study whose corpus composition affects the result.

The field schema and projection portfolio are replay inputs fixed before collision counting.  The portfolio is chosen because the admitted public surfaces expose keyword/operator labels and a smaller yes/no control family; it does not exhaust all possible optimizer vocabularies.  Layer plus placement is reported as an interpretable separating basis for headline witnesses after enumerating retained-field vocabularies; the frontier tables give scoped safety and do not claim global safety.  Public source records admit claims; EXPLAIN, profile, or stage surfaces bind exposed plan evidence when available; SQL execution validates SQL constructability and feature representation rather than private optimizer truth.  A counted witness requires two source-supported reference signatures, projection equality under the declared vocabulary, and separation by the reported field.

The denominator dashboard in Table~\ref{tab:denominator-dashboard} checks that this conservatism does not create unreachable contracts or syntactic probes.  A finite-disjunction guard semantics evaluates each admitted rule against the generated probe denominator.  All 287 grounded rules have at least one triggering probe; 252 depend on query features rather than global engine labels; and no guard uses a feature dimension or value outside the probe-suite domain.

\input{tables/tab_guard_quality}

Coverage is reported over two denominators.  The first is probe-suite coverage: every targeted renderable valid feature interaction is covered by at least one generated probe.  The second is contract triggering: every grounded rule is applicable to at least one probe.  Together these denominators make the empirical scope explicit.  A rule that cannot be triggered by any probe would indicate a probe-suite gap; a feature interaction that cannot be rendered, lacks a source-supported optimizer action, or falls outside the declared feature domain is excluded from the coverage denominator and reported as a modeling constraint.  Thus ``90/90'' below means coverage of declared optimizer-decision requirements inside the feature schema, not coverage of all behavior present in JOB, TPC-H, TPC-DS, or any vendor engine.

The benchmark also exposes executable SQL shapes.  Each generated probe has a normalized SQL skeleton and shape checks for joins, filters, grouping, ordering, limits, and reuse structure.  The full 4,216-probe corpus is accompanied by a compact 48-probe motif-cover prefix.  The real-engine motif execution set uses the deterministic union of this motif cover and residual feature-axis maxima; it is not a hand-picked success set, and it is not a replacement for the full evaluation.

A common failure mode in generated probe suites is to produce feature-complete but non-executable query text.  OptSemBench-C therefore validates every SQL skeleton on a deterministic relational catalog with local, remote, and cross-source namespaces.  This validation is not a performance baseline, contract-truth check, or exhaustive trigger proof.  Its role is stricter and more basic: every generated probe must be parseable, plannable, executable, and shape-consistent with its feature vector.  Table~\ref{tab:denominator-dashboard} reports the validation result: all 4,216 generated probes obtain a plan and execute, with zero SQL-shape failures.  Repeating the full corpus across three deterministic catalog populations yields 12,648 successful probe-catalog executions, and DuckDB/PostgreSQL checks add 8,432 full-corpus engine--probe executions plus 142 motif-representative executions with zero execution failures.

Contract validation is layered.  Guard reachability says that each public rule has a generated probe satisfying its admitted guard; SQL execution says that the denominator is not syntactic fiction.  Visible-plan sanity checks on DuckDB and PostgreSQL expose expected plan tokens at mean rates 0.705 and 0.796 over the full corpus, and 0.756 and 0.819 on feature-space representatives.  The missing 20--30\% is a reported observability limit: joins match completely, while aggregate and CTE labels are often absent from exposed plan text.  These rows bound generated-SQL quality and observability; they do not validate closed-source contracts, reverse-engineer private optimizers, prove every documented behavior fires on every cost-model instance, or cross-validate a source by running the same engine.  A counted witness still requires source-supported reference signatures, guard reachability, projection equality, and field separation; runtime plan satisfaction is a post-admission SQL-denominator validation layer and is excluded by a non-interference check from collision tables, while executability constraints still delimit which feature interactions enter the probe suite.

The same fixed probes support two budgeted evaluation orders.  The cover order is the generator's feature-interaction order.  The diagnostic order greedily maximizes newly triggered grounded rules without adding any new probe; it is a rule-coverage schedule over admitted contracts, not an unseen workload sample.  Table~\ref{tab:benchmark-efficiency} reports both.  The cover order triggers all grounded rules within 100 probes; the diagnostic order triggers 286 of 287 rules within 50 probes and all rules within 100.  Random subsets of 100 probes trigger 81\% of rules on average and 85\% at the 95th percentile.  Thus OptSemBench-C is not merely large enough to cover the surface; it also contains compact diagnostic prefixes that expose almost all grounded contracts.

\input{tables/tab_benchmark_efficiency}

A probe suite for public optimizer contracts must connect to existing database workloads without becoming another latency benchmark.  We therefore crosswalk published optimizer and benchmark motifs to OptSemBench-C feature requirements: JOB join-order sensitivity~\cite{leis2015good}, POLAR-style join structure~\cite{justen2024polar}, TPC decision-support benchmarking~\cite{poess2000tpc}, TPC-DS design~\cite{nambiar2006tpcds}, SSB~\cite{oneil2009ssb}, LDBC~\cite{erling2015ldbc}, Wisconsin~\cite{bitton1983wisconsin}, warehouse surveys~\cite{darmont2007benchmarking}, YCSB~\cite{cooper2010ycsb}, and LakeBench~\cite{deng2024lakebench}.  Figure~\ref{fig:external-motifs} reports a feature-space crosswalk, not workload substitution: 90 declared optimizer-decision requirements across 12 published families are represented by 71 distinct generated SQL probes because several probes satisfy more than one requirement.  Five anchors illustrate the span: JOB join-order search (8/8 requirements), exact-cardinality testing (9/9), CardEst-style plan-quality sensitivity (8/8), adaptive replanning (9/9), and cross-source joinability (7/7).  Table~\ref{tab:motif-difficulty} gives the 12-family matching-probe distribution.  The selected probes are feature-interaction representatives, not replacements for JOB or TPC-H queries, and the check is schema-adjacent rather than independent workload validation.  A counted collision still requires two source-supported signatures, projection equality, side-balanced support, and field separation.  We claim contract-denominator adequacy and feature-space representation, not workload reproduction.

\input{figures/fig_external_motifs_plot}
\input{tables/tab_motif_difficulty}

The feature-space difficulty view prevents a misleading denominator.  Some requirements are broad and match many probes, such as star-schema group-by shapes.  Others are sparse, such as a TPC-H conjunctive-filter requirement that has one matching rendered probe under the current validity constraints.  Reporting both coverage and matching-probe counts makes probe-suite adequacy inspectable: coverage says the requirement is present, while difficulty says how concentrated or redundant that coverage is.  The executable crosswalk adds a second denominator check: for each published requirement, the system selects one satisfying probe and validates the generated SQL on the deterministic catalog, yielding 90 validated requirements, 71 distinct representatives, and 383 result rows.  Thus the denominator is not a list of names; it is a set of optimizer-decision requirements with concrete SQL representatives and an explicit many-to-one map.  Engine-plan validation is a separate experiment because the crosswalk is a denominator check rather than a claim that published benchmark queries and generated probes induce identical plans.
