\section{Grounded Case Studies}
\label{sec:cases}
The aggregate results are explained by four grounded case families, summarized in Table~\ref{tab:cases}.  Each family isolates the optimizer-semantics field that repairs false portability in the aggregate analysis.

\input{tables/tab_cases}
\FloatBarrier

\textbf{Pushdown is an operator family.} Predicate movement, aggregation delegation, join delegation, limit delegation, and Top-N delegation are separate contracts.  A feature matrix erases the operator and placement distinctions that determine whether computation remains inside the engine or moves to a source.  This is why keyword projection fails even though the word ``pushdown'' is common across system descriptions: the semantic object is not the word but the guarded action.  The repair certificate identifies placement because local rewrite and source delegation are the dominant false-equivalence separator.

\textbf{Plan evidence is not isomorphic.} Logical plans, estimated physical plans, runtime profiles, and distributed stage graphs are different public evidence states.  Native operator-name comparison treats them as interchangeable; OptSem-C does not.  This distinction matters for systems whose public interfaces expose logical plans, estimated local operators, cloud-service stage graphs, or runtime reoptimization evidence.  The framework represents these as observability and decision-time fields rather than as formatting differences.

\textbf{Decision time changes semantics.} Compile-time join selection, stage-boundary repartitioning, and runtime adaptive execution are not variants of the same behavior.  Removing decision time makes adaptive and static contracts appear portable.  This connects directly to adaptive query-processing work: a public runtime-adaptive contract is stronger than a compile-time-only public plan even if both systems ultimately produce a correct SQL result \cite{pavlo2017selfdriving,kraska2018learnedindex,krishnan2018sage,yang2022bao,dutt2019selectivity,naru2019}.

\textbf{Materialization and reuse are guarded.} CTE reuse, volatility, grouped aggregation, and explicit materialization controls change whether an intermediate is folded, materialized, or blocks pushdown.  These are query-guarded contract differences, not engine-level features.  A benchmark that does not include reuse and materialization guards can miss precisely the contracts that decide whether two systems are comparable for multi-use analytical subplans.

The cases also explain why the repair fields in Table~\ref{tab:field-repair} are not interchangeable.  In the pushdown case, the repair is operator and placement: a predicate moved inside the engine and a join delegated to a source are different actions even if both use the word pushdown.  In the observability case, the repair is observability and decision time: a logical plan, an estimated physical plan, and a runtime stage graph are different evidence layers.  In the materialization case, the repair is guard and variant: reuse, volatility, and aggregation change whether folding, materialization, or pushdown blocking applies.

These cases serve the role that microbenchmarks serve in a runtime paper: they explain aggregate results by isolating mechanisms.  The pushdown case explains why feature names collide.  The observability case explains why native plan comparison fails.  The decision-time case explains why runtime-adaptive behavior cannot be compared to compile-time planning.  The materialization case explains why query guards, rather than engine labels, determine which contract applies.
