\section{Implications and Related Work}
\label{sec:discussion}
\textbf{Optimizer interfaces as semantic interfaces.}  The results suggest a reporting rule for optimizer studies: a public claim should name the action being compared, the guard under which it applies, the optimizer layer that owns it, the placement at which it executes, the decision time, and the evidence a user can observe.  These fields do not reveal proprietary cost formulas.  They specify the public boundary conditions needed to decide whether two engines are being compared on the same query-processing phenomenon.  In heterogeneous analytics, where optimizers span local engines, adapters, remote sources, and service-side stages \cite{haas1997garlic,stonebraker2015bigdawg,calcite_docs,datafusion_docs,presto_docs}, the distinction is decisive: a speedup or plan difference may reflect local rewrite, source delegation, exchange placement, runtime adaptation, or merely a different evidence surface.

\textbf{Relation to optimizer theory and testing.}  Classical optimizers separate logical equivalence, physical alternatives, access paths, search, costing, and runtime adaptation \cite{selinger1979access,graefe1990volcano,graefe1993volcano,graefe1995cascades,chaudhuri1998overview,ioannidis1996query}.  OptSem-C does not reconstruct private search.  It asks which of those distinctions are public enough to support cross-engine comparison.  The recovered repair fields are therefore architectural: placement separates local execution from source or service execution; layer separates rewrite, physical planning, distributed planning, and runtime adaptation; decision time separates compile-time choice from mid-query change; observability separates estimated plans from runtime evidence.  SQL-equivalence systems decide whether expressions preserve results \cite{chu2017cosette,chu2018usemiring,wang2024qed,chandra1977optimal,aho1979equivalences}; OptSem-C studies a different relation in which result-equivalent queries can expose non-equivalent optimizer contracts.  Provenance work supplies the methodological precedent that support should be attached to the object being compared \cite{green2007provenance,buneman2001why,cheney2009provenance,green2010provenance}.  Exact-cardinality testing, JOB, adaptive processing, and learned-optimizer work similarly sharpened denominators when older metrics hid the behavior of interest \cite{leis2015good,leis2018looking,chaudhuri2009exact,markl2004progressive,kabra1998reoptimization,deshpande2007adaptive,avnur2000eddies,marcus2019neo,marcus2021bao,zhu2023lero}.

\textbf{Relation to benchmarking.}  TPC-H, TPC-DS, SSB, LDBC, Wisconsin, AS3AP, LinkBench, ClickBench, CardBench, and LakeBench matter because they make workloads, labels, or evaluation denominators explicit \cite{tpcds,tpch,oneil2009ssb,erling2015ldbc,bitton1983wisconsin,turbyfill1989as3ap,armstrong2013linkbench,clickbench,chronis2024cardbench,deng2024lakebench}.  OptSemBench-C borrows that discipline without becoming another latency benchmark.  Its unit is a query probe over an optimizer-decision feature interaction.  A thousand constant variants can leave the contract surface unchanged, whereas one query that crosses a source boundary, introduces reuse, changes a join shape, or triggers adaptivity can reveal a new public behavior.  This is why the evaluation reports feature-interaction coverage, contract triggering, external optimizer motifs, executable SQL validation, semantic repair, witness dispersion, and leave-out stability rather than a single workload score.

\textbf{What changes for comparative studies.}  A future optimizer comparison can use the semantics as a front-end calibration step.  First state the public contract being compared; then state the guard and semantic fields that make the behavior comparable.  If exact signatures agree under those fields, subsequent performance or plan-quality measurements have a clear denominator.  If a coarse projection collapses unequal signatures, the comparison can still report the measurement, but the portability claim must name the missing field.  The field lattice matters because the answer is neither ``all details'' nor ``any feature label.''  In the no-variant semantic universe, every singleton field is unsafe, while two-field semantic vocabularies can already reach the exact-equivalence denominator.  The dangerous region is under-specified heterogeneity, not heterogeneity itself.

\textbf{Scope of the contribution.}  The paper sits between foundations and systems.  The formal contribution is guarded modal contract semantics, projection-kernel characterization, and repair-frontier certificates.  The systems contribution is the construction of a public optimizer-contract relation over analytical engines and executable query probes.  The benchmark contribution is the feature-interaction denominator and its connection to established workload motifs.  The combination is necessary: without formal semantics, the corpus is a feature audit; without grounding, the theory is only a toy counterexample; without executable probes, the benchmark denominator is not testable.  OptSem-C ties those pieces together so that optimizer portability can be stated, falsified, and repaired at the same semantic resolution.

\section{Conclusion}
A coarse optimizer comparison can fabricate portability even when every individual system statement is public and accurate.  OptSem-C formalizes this failure as a projection kernel over guarded public optimizer contracts.  The paper defines the contract object, gives a finite state-join and projection semantics, generates covering query probes, grounds contracts in public evidence, executes the generated probes, and computes the semantic fields that repair observed false equivalences.

Keyword projection declares 2,284 equivalences, 254 of which exact contracts refute; operator-only projection declares 2,268 equivalences, 238 of which are false.  The false pairs are not near misses: the headline witnesses have disjoint exact public-contract signatures.  The repair is semantic rather than lexical.  Placement repairs keyword collisions, layer repairs operator-only collisions, and layer plus placement repairs all 498 headline witnesses when fine-grained action-variant labels are excluded.  The exhaustive retained-field lattice strengthens the point: over the no-variant semantic universe, every singleton field is unsafe, but two-field semantic vocabularies already reach the exact-equivalence denominator.  A comparison that preserves the public contract surface can still disagree on speed, cost, or plan quality; a comparison that does not preserve it has already changed the object being measured.


The broader consequence is that optimizer portability should be treated as a typed relation rather than a scalar capability.  A relation has a domain, a guard, an observable surface, and an equivalence criterion.  That view explains why the same SQL text can support a valid latency comparison and an invalid portability claim at the same time: the query may run on both systems while one public contract concerns connector delegation and the other concerns local physical planning.  OptSem-C makes this distinction explicit before measurements are interpreted.

The repair frontier gives benchmark authors a compact protocol.  Report the workload as usual, but attach every optimizer-portability claim to the contract signature it uses.  If the signature is projected, the projection should carry a repair certificate or a counter-witness.  New engines, actions, and workload motifs then refine the same contract relation rather than replacing an informal feature table.  Coarse optimizer matrices are therefore not unusable; they are incomplete.  OptSem-C gives a finite test for incompleteness and a finite repair when the test fails.  This is the intended unit of progress: each additional public engine, source class, or workload motif can be absorbed by the same guarded relation and then checked against the same kernel and frontier.
